### 2022-01-19 proposal.Rmd
added proposal

---
title: "Project Proposal"
author: "Peyton Kuhlers"
date: "1/18/2022"
output: pdf_document
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Patient Derived Xenografts (PDX) are *in vivo* cancer models
in which patient-derived tumors are taken and engrafted into immuno-deficient mice.
Once established, PDXs can be used as platforms for drug discovery or
investigation of *in vivo* drug mechanisms.

The PDX Development and Trial Centers Research Network (PDXNet) is an
NCI consortium that has generated publicly available for
PDX models across a large number of cancer types.
This data includes genomic characterizations and tumor growth measurements for
a variety of drugs.

My goal is to analyze the drug response data of one or two
cancers to determine if any of the drugs have a significant effect 
on the growth of tumors in PDX models.
Growth data for colon cancer includes ~2700 volume measurements,
7 drugs (and 1 control), and 263 mice originating from 7 unique patients.

Because there is also a wealth of genomic data available as well,
I may incorporate data such as tumor mutation or subtype --if feasible--
to identify any mutation or subtype specific drug responses.

### 2022-02-02 proposal.Rmd
updated proposal

@@
 author: "Peyton Kuhlers"
-date: "1/18/2022"
 output: pdf_document
+geometry: "left=2cm,right=2cm,top=0.25cm,bottom=1.5cm"
 fontsize: 12pt
@@
 Once established, PDXs can be used as platforms for drug discovery or
-investigation of *in vivo* drug mechanisms.
-
+investigation of *in vivo* drug mechanisms (1).
 The PDX Development and Trial Centers Research Network (PDXNet) is an
@@
 
-My goal is to analyze the drug response data of one or two
-cancers to determine if any of the drugs have a significant effect 
-on the growth of tumors in PDX models.
+Evrard et al (2) used multiple strategies to evaluate tumor response
+from three PDX models treated with temozolomide.
+These included percent change in tumor volume,
+area under the curve, and progression free survival, among others,
+but did not use any linear models of the volume measurements directly.
+Pearson et al (3) characterized the growth rates of PDX tumors
+across passages using linear mixed models and noted that
+tumor volumes on the log-scale seem to grow quadratically.
+
+My proposal is to analyze the drug response data
+available from the PDXnet for one or two
+cancers to determine if any of the tested drugs significantly inhibit
+tumor growth.
+Data will be downloaded from the PDXNet online portal
+(https://portal.pdxnetwork.org/).
 Growth data for colon cancer includes ~2700 volume measurements,
 7 drugs (and 1 control), and 263 mice originating from 7 unique patients.
+I plan to use a linear model fit with generalized least squares
+to determine if any treatments slow tumor growth.
+Additionally, I plan to use a mixed model approach to
+investigate the variation in response to treatment.
 
-Because there is also a wealth of genomic data available as well,
-I may incorporate data such as tumor mutation or subtype --if feasible--
-to identify any mutation or subtype specific drug responses.
+The PDXNet portal also contains genomic data,
+such as tumor mutational burden (TMB)
+and homologous recombination defficiency analysis (HRD),
+which may impact response to treatment (4,5).
+I aim to incorporate TMB and HRD measurements into the linear models and
+test for their effect on tumor growth and interaction with treatment.
+
+References
+----------
+1) Tentler, J., Tan, A., Weekes, C. et al. Patient-derived tumour xenografts as models for oncology drug development. Nat Rev Clin Oncol 9, 338–350 (2012). https://doi.org/10.1038/nrclinonc.2012.61
+2) Evrard, Yvonne A et al.
+“Systematic Establishment of Robustness and Standards in Patient-Derived Xenograft Experiments and Analysis.” Cancer research vol.
+80,11 (2020): 2286-2297.
+doi:10.1158/0008-5472.CAN-19-3101
+3) Pearson, Alexander T., et al. "Patient-derived xenograft (PDX) tumors increase growth rate with time." Oncotarget [Online], 7.7 (2016): 7993-8005. Web. 2 Feb. 2022 
+4) Yarchoan, Mark et al. “Tumor Mutational Burden and Response Rate to PD-1 Inhibition.” The New England journal of medicine vol. 377,25 (2017): 2500-2501. doi:10.1056/NEJMc1713444
+5) Aguirre, Andrew J et al.
+“Real-time Genomic Characterization of Advanced Pancreatic Cancer to Enable Precision Medicine.” Cancer discovery vol.
+8,9 (2018): 1096-1111.
+doi:10.1158/2159-8290.CD-18-0275

### 2022-03-01 src/plot_growth.R
plot growth curves

@@ -1,0 +1,13 @@
+## Plot tumor volume over time by treatment
+## Replicate plots from pdxnet portal
+library(ggplot2)
+vol <- read.csv("source_data/pdx_colon.csv")
+theme_set(theme_classic())
+
+pdf("figures/growth_plots.pdf")
+ggplot(vol, aes(x = OBS_DAY, y = TUMOR_WT, color = ID)) +
+  geom_line(alpha = 0.5) +
+  facet_wrap(~ AgentName) +
+  guides(color = "none") +
+  labs(x = "Observation Day", y = expression(paste("Tumor Volume (m", m^3, ")")))
+dev.off()

### 2022-03-01 proposal_presentation.Rmd
proposal presentation

---
title: "Proposal Presentation"
author: "Peyton Kuhlers"
date: "3/2/2022"
output: 
  beamer_presentation:
    theme: "AnnArbor"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
## Background

- Patient Derived Xenografts (PDX) are *in vivo* cancer models used as platforms
for drug discovery, biomarker identification, and mechanistic studies\footnote{Tentler et al 2012}

\centering
\includegraphics[width = 8cm, height = 4cm]{figures/pdx_graphic.png}

\note{Both in PDXnet and in the literature in general}

## Background Cont.

- The PDXNet is an NCI-sponsored consortium aimed at using PDX models
for high-throughput development and testing of cancer therapies

- Much work has centered around genomic characterization of PDX models\footnote{Sun et al 2021}\textsuperscript{,}\footnote{Woo et al 2020}

- While data is available, comparatively less work has focused on 
analysis of tumor growth data in PDX models.

## Previous Work

- PDXNet collaborators\footnote{Evrard et al 2020} tested multiple
different strategies for analyzing tumor growth under treatment
  + Percent change
  + Modified RECIST
  + Area Under the Growth Curve
  + "Tumor growth inhibition"

\note{Lots of ignored data, not modelling the entire time course}
\note{Plus multiple clusters, mouse, patient, etc}

- In contrast, Pearson\footnote{Pearson et al 2016} and Oberg\footnote{Oberg et al 2021}
used mixed models to analyze growth over multiple passages and treatment effects, respectively.

## Proposal

- Given PDXs are meant for high-throughput discovery,
robust and **efficient** methods are necessary
  + Inefficient statistical methods waste resources

- My proposal is to used a mixed model approach to analyze the PDXNet growth data

## Data

- Growth data for colon cancer includes ~2700 volume measurements,
7 drugs (and 1 control), and 263 mice originating from 7 unique patients.

\centering
\includegraphics[width = 9cm, height = 5.5cm]{figures/growth_plots.pdf}

## Model

### 2022-03-01 driver.sh
driver script

@@ -1,0 +1,3 @@
+## Driver script to automate analysis and reporting
+
+Rscript src/plot_growth.R

### 2022-03-02 proposal_presentation.Rmd
presentation update

@@
 ---
-title: "Proposal Presentation"
+title: "Analysis of PDX Growth Data"
 author: "Peyton Kuhlers"
@@
 ```
+
 ## Background
@@
 
+- Becoming preferred model as they preserve donor genomic characteristics
+and allow for microenvironment modelling *in vivo*
+
 \centering
@@
 
-\note{Both in PDXnet and in the literature in general}
+\note{In contrast to cells in a dish}
 
@@
 analysis of tumor growth data in PDX models.
+
+\note{Both in PDXnet and in the literature in general}
 
@@
 
-## Proposal
+## Significance
 
@@
   + Inefficient statistical methods waste resources
-
-- My proposal is to used a mixed model approach to analyze the PDXNet growth data
 
@@
 
-## Model
+## Analysis
+
+- Mixed effects model of tumor growth
+
+\centering
+$E[Y_{i}|b_i] = \beta_0 + b_{i1} + \beta_1(time_{i}) + b_{i2}(time) + \gamma_i (agent_i*time)$
+
+## Result Reporting
+
+- Profile Plots (shown above)
+
+- Forest plot of interaction coefficients
+
+- Table of specific pairwise contrasts between different therapies
+
+- Variance of random effects
+
+- As a secondary analysis, test methods from Evrard et al 
+at e.g. day 75 and compare to model contrasts at day 75.
+
+##
+
+Thank You! Questions?

### 2022-03-02 src/plot_growth.R

@@ -5,9 +5,10 @@ vol <- read.csv("source_data/pdx_colon.csv")
 theme_set(theme_classic())
 
 pdf("figures/growth_plots.pdf")
-ggplot(vol, aes(x = OBS_DAY, y = TUMOR_WT, color = ID)) +
-  geom_line(alpha = 0.5) +
+ggplot(vol, aes(x = OBS_DAY, y = log(TUMOR_WT))) +
+  geom_line(alpha = 0.5, aes(group = ID)) +
+  geom_smooth(method = 'lm', formula = y~x, se = F) +
   facet_wrap(~ AgentName) +
   guides(color = "none") +
-  labs(x = "Observation Day", y = expression(paste("Tumor Volume (m", m^3, ")")))
+  labs(x = "Observation Day", y = expression(paste("Log Tumor Volume (m", m^3, ")")))
 dev.off()

### 2022-04-06 paper_Kuhlers.Rmd
starting paper

---
title: "Analysis of Patient Derived Xenograft Growth Curves"
author: "Peyton Kuhlers"
date: "4/6/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
```

Introduction
=============

Patient Derived Xenografts (PDX) are *in vivo* cancer models
in which patient tumors are removed and engrafted into immuno-deficient mice.
Once established, PDXs can be used as platforms for drug discovery or
investigation of *in vivo* drug mechanisms (**REFS**).

PDX models have significant advantages over alternative models.
Standard *in vitro* models typically undergo transformations
to ensure proper growth and are kept in growth media that is drastically
different from the tumor enviroment *in vivo*. Additionally, these models
are "two dimensional" and all three dimensional interactions are
lost with cell lines (**REFS**).
PDX models preserve the 3D nature of human tumors and do not require
transformation, thereby also preserving the genomic characteristics of the
donor tissue (**REFS**).

Methods
=======

Tumor volume data was obtained from the PDXnet Portal,
queried for all colon adenocarcinoma.

Results
=======

\begin{centering}
\includegraphics[width = 6in, height = 8in]{figures/growth_plots.pdf}
\end{centering}

Discussion
===========

### 2022-04-06 src/mixed.R
mixed models

@@ -1,0 +1,44 @@
+## Analysis of colon PDX growth using mixed models
+library(tidyverse)
+library(lme4)
+growth <- read.csv("source_data/pdx_colon.csv") %>%
+  mutate(day = OBS_DAY - 7,
+         drug = relevel(factor(AgentName), ref = "Control"))
+
+## OLS estimates
+fit0 <- glm(
+  TUMOR_WT ~ OBS_DAY + OBS_DAY:drug,
+  data = growth,
+  family = gaussian(link = "log"))
+summary(fit0)
+
+## Random Intercept by mouse
+fit1 <- glmer(
+  TUMOR_WT ~ OBS_DAY + OBS_DAY:drug + (1|ID),
+  data = growth,
+  family = gaussian(link = "log")
+)
+summary(fit1)
+
+## Random slope and intercept by mouse
+fit2 <- glmer(
+  TUMOR_WT ~ OBS_DAY + OBS_DAY:drug + (1 + OBS_DAY|ID),
+  data = growth,
+  family = gaussian(link = "log"), 
+)
+summary(fit2)
+
+## Exploring some different model fitting methods
+library(nlme)
+library(MASS)
+fit <- lme(TUMOR_WT ~ OBS_DAY + OBS_DAY:drug, random = ~ OBS_DAY | ID, data = growth2)
+summary(fit)
+fit <-
+  MASS::glmmPQL(
+    TUMOR_WT ~ OBS_DAY + OBS_DAY:drug,
+    random = ~ OBS_DAY |
+      ID,
+    family = gaussian(link = "log"),
+    data = growth
+  )
+summary(fit)

### 2022-04-06 src/models.R

@@ -1,21 +1,0 @@
-## Analysis of colon PDX growth using mixed models
-library(tidyverse)
-library(lme4)
-growth <- read.csv("source_data/pdx_colon.csv") %>%
-  mutate(day = OBS_DAY - 7)
-
-## Random Intercept by mouse
-fit1 <- glmer(
-  TUMOR_WT ~ OBS_DAY + OBS_DAY:AgentName + (1|ID),
-  data = growth,
-  family = gaussian(link = "log")
-)
-summary(fit1)
-
-## Random slope and intercept by mouse
-fit2 <- glmer(
-  TUMOR_WT ~ OBS_DAY + OBS_DAY:AgentName + (1 + OBS_DAY|ID),
-  data = growth,
-  family = gaussian(link = "log")
-)
-summary(fit2)

### 2022-04-06 src/aft.R
exploring aft models

@@ -1,0 +1,17 @@
+## Analysis of colon PDX growth using mixed models
+library(tidyverse)
+library(survival)
+growth <- read.csv("source_data/pdx_colon.csv") %>%
+  mutate(day = OBS_DAY - 7,
+         drug = relevel(factor(AgentName), ref = "Control"),
+         censor = 1)
+gg <- growth %>%
+  group_by(ID, drug) %>%
+  summarize(maxDay = max(OBS_DAY)) %>%
+  mutate(censor = 1)
+aft <- survreg(Surv(maxDay, censor) ~ drug, data = gg)
+
+growth <- read.csv("source_data/pdx_pancreas.csv") %>%
+  mutate(day = OBS_DAY - 7,
+         drug = relevel(factor(AgentName), ref = "Control"),
+         censor = 1)

### 2022-04-06 src/aft.R
testing diff models

@@ -3,15 +3,11 @@ library(tidyverse)
 library(survival)
 growth <- read.csv("source_data/pdx_colon.csv") %>%
   mutate(day = OBS_DAY - 7,
-         drug = relevel(factor(AgentName), ref = "Control"),
-         censor = 1)
+         drug = relevel(factor(AgentName), ref = "Control"))
 gg <- growth %>%
   group_by(ID, drug) %>%
   summarize(maxDay = max(OBS_DAY)) %>%
   mutate(censor = 1)
 aft <- survreg(Surv(maxDay, censor) ~ drug, data = gg)
-
-growth <- read.csv("source_data/pdx_pancreas.csv") %>%
-  mutate(day = OBS_DAY - 7,
-         drug = relevel(factor(AgentName), ref = "Control"),
-         censor = 1)
+summary(aft)
+saveRDS(aft, "derived_data/aft_model.rds")

### 2022-04-06 src/mixed.R

@@ -3,6 +3,7 @@ library(tidyverse)
 library(lme4)
 growth <- read.csv("source_data/pdx_colon.csv") %>%
   mutate(day = OBS_DAY - 7,
+         month = OBS_DAY / 30,
          drug = relevel(factor(AgentName), ref = "Control"))
 
 ## OLS estimates
@@ -24,16 +25,23 @@ summary(fit1)
 fit2 <- glmer(
   TUMOR_WT ~ OBS_DAY + OBS_DAY:drug + (1 + OBS_DAY|ID),
   data = growth,
-  family = gaussian(link = "log"), 
+  family = gaussian(link = "log")
 )
 summary(fit2)
+
+## Random slope and intercept by mouse -- by month
+fitM <- glmer(
+  TUMOR_WT ~ month + month:drug + (1 + month|ID),
+  data = growth,
+  family = gaussian(link = "log")
+)
+saveRDS(fitM, "derived_data/colon_month_glmer.rds")
 
 ## Exploring some different model fitting methods
 library(nlme)
 library(MASS)
-fit <- lme(TUMOR_WT ~ OBS_DAY + OBS_DAY:drug, random = ~ OBS_DAY | ID, data = growth2)
-summary(fit)
-fit <-
+## log-linear model fit with PQL
+fit_day_pql <-
   MASS::glmmPQL(
     TUMOR_WT ~ OBS_DAY + OBS_DAY:drug,
     random = ~ OBS_DAY |
@@ -41,4 +49,15 @@ fit <-
     family = gaussian(link = "log"),
     data = growth
   )
-summary(fit)
+summary(fit_day_pql)
+
+fit_month_pql <-
+  MASS::glmmPQL(
+    TUMOR_WT ~ month + month:drug,
+    random = ~ month |
+      ID,
+    family = gaussian(link = "log"),
+    data = growth
+  )
+summary(fit_month_pql)
+saveRDS(fit, "derived_data/colon_kk")

### 2022-04-06 src/plot_effects.R
plotting model estimates

@@ -1,0 +1,4 @@
+library(sjPlot)
+aft <- readRDS("derived_data/aft_model.rds")
+
+plot_model(aft, title = "AFT Model Acceleration Factors")

### 2022-04-19 src/clean.R
added cleaning script

@@ -1,0 +1,19 @@
+library(tidyverse)
+growth <- read.csv("source_data/pdx_colon.csv") %>%
+  mutate(
+    day = OBS_DAY - 7,
+    month = OBS_DAY / 30,
+    drug = relevel(factor(AgentName), ref = "Control")
+  ) %>%
+  group_by(ID) %>%
+  mutate(
+    first = min(OBS_DAY),
+    maxVol = max(TUMOR_WT),
+    days_since_first = OBS_DAY - first,
+    reached_1500 = ifelse(maxVol >= 1500, 1, 0),
+    total_obs = length(OBS_DAY),
+    days_to_1500 = days_since_first[which.min(1500 - TUMOR_WT)]
+  ) %>%
+  ungroup()
+
+write.csv(growth, "derived_data/pdx_colon_clean.csv")

### 2022-04-19 src/aft.R
naive aft

@@ -1,13 +1,18 @@
 ## Analysis of colon PDX growth using mixed models
 library(tidyverse)
 library(survival)
-growth <- read.csv("source_data/pdx_colon.csv") %>%
-  mutate(day = OBS_DAY - 7,
-         drug = relevel(factor(AgentName), ref = "Control"))
-gg <- growth %>%
-  group_by(ID, drug) %>%
-  summarize(maxDay = max(OBS_DAY)) %>%
-  mutate(censor = 1)
-aft <- survreg(Surv(maxDay, censor) ~ drug, data = gg)
+growth <- read.csv("derived_data/pdx_colon_clean.csv") %>%
+  mutate(drug = relevel(factor(AgentName), ref = "Control")) %>%
+  filter(total_obs > 1)
+
+## AFT model for days to 1500 (or next closest day)
+aft <- survreg(Surv(days_to_1500, reached_1500) ~ drug, data = growth)
 summary(aft)
 saveRDS(aft, "derived_data/aft_model.rds")
+
+ggsurvplot(
+  survfit(Surv(days_to_1500, reached_1500) ~ drug, data = growth),
+  linetype = c("dashed", rep("solid", 7)),
+  censor.shape = "|"
+)
+

### 2022-04-19 src/mixed.R
more mixed model exploration

@@ -1,33 +1,33 @@
 ## Analysis of colon PDX growth using mixed models
 library(tidyverse)
 library(lme4)
-growth <- read.csv("source_data/pdx_colon.csv") %>%
-  mutate(day = OBS_DAY - 7,
-         month = OBS_DAY / 30,
-         drug = relevel(factor(AgentName), ref = "Control"))
+growth <- read.csv("derived_data/pdx_colon_clean.csv") %>%
+  mutate(drug = relevel(factor(AgentName), ref = "Control"),
+         months_since_first = days_since_first/30)
 
 ## OLS estimates
 fit0 <- glm(
-  TUMOR_WT ~ OBS_DAY + OBS_DAY:drug,
+  TUMOR_WT ~ month + month:drug,
   data = growth,
   family = gaussian(link = "log"))
 summary(fit0)
 
-## Random Intercept by mouse
-fit1 <- glmer(
-  TUMOR_WT ~ OBS_DAY + OBS_DAY:drug + (1|ID),
-  data = growth,
-  family = gaussian(link = "log")
-)
-summary(fit1)
-
-## Random slope and intercept by mouse
-fit2 <- glmer(
-  TUMOR_WT ~ OBS_DAY + OBS_DAY:drug + (1 + OBS_DAY|ID),
-  data = growth,
-  family = gaussian(link = "log")
-)
-summary(fit2)
+# ## Random Intercept by mouse
+# fit1 <- glmer(
+#   TUMOR_WT ~ OBS_DAY + OBS_DAY:drug + (1|ID),
+#   data = growth,
+#   family = gaussian(link = "log")
+# )
+# summary(fit1)
+#
+# ## Random slope and intercept by mouse
+# fit2 <- glmer(
+#   TUMOR_WT ~ OBS_DAY + OBS_DAY:drug + (1 + OBS_DAY|ID),
+#   data = growth,
+#   family = gaussian(link = "log"),
+#   glmerControl(optimizer = "bobyqa")
+# )
+# summary(fit2)
 
 ## Random slope and intercept by mouse -- by month
 fitM <- glmer(
@@ -35,7 +35,23 @@ fitM <- glmer(
   data = growth,
   family = gaussian(link = "log")
 )
+summary(fitM)
 saveRDS(fitM, "derived_data/colon_month_glmer.rds")
+
+fit_first <- glmer(
+  TUMOR_WT ~ months_since_first + months_since_first:drug + (1 + months_since_first|ID),
+  data = growth,
[…]

### 2022-04-19 src/plot_growth.R
growth visualization exploration

@@ -1,9 +1,10 @@
 ## Plot tumor volume over time by treatment
 ## Replicate plots from pdxnet portal
-library(ggplot2)
-vol <- read.csv("source_data/pdx_colon.csv")
+library(tidyverse)
+vol <- read.csv("derived_data/pdx_colon_clean.csv", row.names = 1)
 theme_set(theme_classic())
 
+## Profile Plots facet by drug
 pdf("figures/growth_plots.pdf")
 ggplot(vol, aes(x = OBS_DAY, y = log(TUMOR_WT))) +
   geom_line(alpha = 0.5, aes(group = ID)) +
@@ -12,3 +13,40 @@ ggplot(vol, aes(x = OBS_DAY, y = log(TUMOR_WT))) +
   guides(color = "none") +
   labs(x = "Observation Day", y = expression(paste("Log Tumor Volume (m", m^3, ")")))
 dev.off()
+
+## ExpName ?
+ggplot(vol, aes(x = days_since_first, y = log(TUMOR_WT), color = ExpNameCSV)) +
+  geom_line(alpha = 0.5, aes(group = ID)) +
+  geom_smooth(method = 'lm', formula = y~x, se = F) +
+  facet_wrap(~ AgentName) +
+  guides(color = "none") +
+  labs(x = "Observation Day", y = expression(paste("Log Tumor Volume (m", m^3, ")")))
+
+## Readjusted time origin facet by drug
+ggplot(vol, aes(x = days_since_first, y = log(TUMOR_WT))) +
+  geom_line(alpha = 0.5, aes(group = ID)) +
+  geom_smooth(method = 'lm', formula = y~x, se = F) +
+  facet_wrap(~ AgentName) +
+  guides(color = "none") +
+  labs(x = "Observation Day", y = expression(paste("Log Tumor Volume (m", m^3, ")")))
+
+## Readjust time origin and slopes for each drug
+ggplot(vol, aes(x = days_since_first, y = log(TUMOR_WT))) +
+  geom_line(alpha = 0.2, aes(group = ID)) +
+  geom_smooth(method = 'lm', formula = y~x, se = F, aes(color = AgentName)) +
+  labs(x = "Observation Day", y = expression(paste("Log Tumor Volume (m", m^3, ")")))
+
+## Maximum volume distribution
+ggplot(vol, aes(x = maxVol)) +
+  geom_histogram(bins = 20, color = "black", fill = "lightblue") +
[…]

### 2022-04-20 driver.sh
moved R scripts to own folder

@@ -1,3 +1,5 @@
 ## Driver script to automate analysis and reporting
 
-Rscript src/plot_growth.R
+Rscript src/R/clean.R
+
+Rscript src/R/plot_growth.R

### 2022-04-20 src/R/aft.R
naive and imputed aft models

@@ -1,18 +1,58 @@
 ## Analysis of colon PDX growth using mixed models
 library(tidyverse)
 library(survival)
-growth <- read.csv("derived_data/pdx_colon_clean.csv") %>%
-  mutate(drug = relevel(factor(AgentName), ref = "Control")) %>%
-  filter(total_obs > 1)
+library(survminer)
+growth <- read.csv("derived_data/pdx_colon_clean.csv", row.names = 1) %>%
+  mutate(drug = relevel(factor(AgentName), ref = "Control"))
 
-## AFT model for days to 1500 (or next closest day)
-aft <- survreg(Surv(days_to_1500, reached_1500) ~ drug, data = growth)
+########################################################################
+## AFT model for days to 1500 (or next closest day) -- 'Naive' Model
+########################################################################
+
+
+aft <- survreg(Surv(days_to_1500, reached_1500) ~ drug, data = growth %>% filter(total_obs > 1))
 summary(aft)
 saveRDS(aft, "derived_data/aft_model.rds")
 
+## Kaplan Meier plot to complement the AFT
 ggsurvplot(
   survfit(Surv(days_to_1500, reached_1500) ~ drug, data = growth),
   linetype = c("dashed", rep("solid", 7)),
   censor.shape = "|"
 )
 
+
+###########################################
+## Mixed model imputed AFT
+###########################################
+
+## Extract the coefficients of the mixed model
+## Create a model matrix for the drugs each *SUBJECT* was assigned too
+## Multiply the coefficent matrix with model matrix to get the subject specific equation
+
+fit_mm <- readRDS("derived_data/colon_month_first_glmer.rds")
+mm_coef <- coef(fit_mm)$ID
+trt_by_subj <- growth %>%
+  select(ID, drug) %>%
+  arrange(ID) %>%
+  distinct()
+mod_mat <- cbind(1, model.matrix(~ drug, trt_by_subj))
+subj_effects <- mm_coef * mod_mat
+
+## Model imputed time to 1500 -- mixed model was on the log scale so log(1500)
+
+mm_impute <-
+  data.frame(imputed_time_to_1500 = (log(1500) - subj_effects[, 1]) / rowSums(subj_effects[, c(2:9)])) %>%
+  rownames_to_column("ID") %>%
+  left_join(x = growth,
+            y = .,
+            by = "ID")
+
[…]

### 2022-04-20 src/R/mixed.R

@@ -38,12 +38,14 @@ fitM <- glmer(
 summary(fitM)
 saveRDS(fitM, "derived_data/colon_month_glmer.rds")
 
+## Random slope and intercept by mouse -- months since first observation
 fit_first <- glmer(
   TUMOR_WT ~ months_since_first + months_since_first:drug + (1 + months_since_first|ID),
   data = growth,
   family = gaussian(link = "log")
 )
 summary(fit_first)
+saveRDS(fit_first, "derived_data/colon_month_first_glmer.rds")
 
 ## Random slope and intercept by mouse nested in experiment -- by month
 fitNest <- glmer(

### 2022-04-20 src/R/plot_effects.R

@@ -1,4 +1,13 @@
+## Forest plots of effects
+library(tidyverse)
 library(sjPlot)
-aft <- readRDS("derived_data/aft_model.rds")
 
-plot_model(aft, title = "AFT Model Acceleration Factors")
+
+aft <- readRDS("derived_data/aft_model.rds")
+plot_model(aft, title = "AFT Model Acceleration Factors",)
+
+mm <- readRDS("derived_data/colon_month_first_glmer.rds")
+plot_model(mm, transform = "exp")
+
+aft_impute <- readRDS("derived_data/colon_imputed_aft.rds")
+plot_model(aft_impute)

### 2022-04-21 paper_Kuhlers.Rmd
paper updates

@@
 
+Mice were randomized to treatment groups and did not begin observation
+until reaching a tumor volume of 150-200mm^3 (Evrard et al).
+
+### Data Processing
+
 Tumor volume data was obtained from the PDXnet Portal,
-queried for all colon adenocarcinoma.
+queried for all available colon adenocarcinoma.
+Observation days were rescaled to the number of days the sample
+was observed, so that the time origin was 0 for all samples.
+For example, if the mouse was staged on observation day 30,
+this became day 0, and all subsequent measurements on this mouse were scaled to
+days since day 30. Days were then rescaled to months.
+
+### Mixed Effects Model
+
+Tumor volumes were analyzed using generalized linear mixed effects models.
+Fixed effects included were time and the time-drug interaction.
+The drug main effect was not included to model equal mean weight
+at baseline amongst all treatment groups.
 

### 2022-04-21 presentation_Kuhlers.Rmd
presentation

--- 2022-03-02 proposal_presentation.Rmd
@@
 ---
-title: "Analysis of PDX Growth Data"
+title: "Model Based Analysis of Patient Derived Xenograft Tumor Volumes"
 author: "Peyton Kuhlers"
-date: "3/2/2022"
-output: 
-  beamer_presentation:
-    theme: "AnnArbor"
+date: "4/25/2022"
+output: beamer_presentation
 ---
@@
 
-\note{Lots of ignored data, not modelling the entire time course}
-\note{Plus multiple clusters, mouse, patient, etc}
-
 - In contrast, Pearson\footnote{Pearson et al 2016} and Oberg\footnote{Oberg et al 2021}
@@
 
+- Mice are randomized to treatments and 'staged' to 150-200mm^3.
+
+- 
 \centering
@@
 \centering
-$E[Y_{i}|b_i] = \beta_0 + b_{i1} + \beta_1(time_{i}) + b_{i2}(time) + \gamma_i (agent_i*time)$
+$log(E[Y_{i}|b_i]) = \beta_0 + b_{i1} + \beta_1(time_{i}) + b_{i2}(time) + \gamma_i (agent_i*time)$
 

### 2022-04-21 src/R/clean.R
removed some duplicate rows

@@ -1,5 +1,5 @@
 library(tidyverse)
-growth <- read.csv("source_data/pdx_colon.csv") %>%
+growth <- read.csv("source_data/pdx_colon.csv", row.names = 1) %>%
   mutate(
     day = OBS_DAY - 7,
     month = OBS_DAY / 30,
@@ -14,6 +14,7 @@ growth <- read.csv("source_data/pdx_colon.csv") %>%
     total_obs = length(OBS_DAY),
     days_to_1500 = days_since_first[which.min(1500 - TUMOR_WT)]
   ) %>%
-  ungroup()
+  ungroup() %>%
+  distinct()
 
 write.csv(growth, "derived_data/pdx_colon_clean.csv")

### 2022-04-21 src/R/aft.R
added model plots

@@ -2,6 +2,8 @@
 library(tidyverse)
 library(survival)
 library(survminer)
+library(sjPlot)
+theme_set(theme_classic())
 growth <- read.csv("derived_data/pdx_colon_clean.csv", row.names = 1) %>%
   mutate(drug = relevel(factor(AgentName), ref = "Control"))
 
@@ -12,15 +14,19 @@ growth <- read.csv("derived_data/pdx_colon_clean.csv", row.names = 1) %>%
 
 aft <- survreg(Surv(days_to_1500, reached_1500) ~ drug, data = growth %>% filter(total_obs > 1))
 summary(aft)
-saveRDS(aft, "derived_data/aft_model.rds")
+
+pdf("figures/naive_aft_forest.pdf")
+plot_model(aft, rm.terms = "Log(scale)", title = "", sort.est = T, axis.title = c("Acceleration Factors", ""))
+dev.off()
 
 ## Kaplan Meier plot to complement the AFT
+pdf("figures/naive_km.pdf")
 ggsurvplot(
   survfit(Surv(days_to_1500, reached_1500) ~ drug, data = growth),
   linetype = c("dashed", rep("solid", 7)),
   censor.shape = "|"
 )
-
+dev.off()
 
 ###########################################
 ## Mixed model imputed AFT
@@ -50,9 +56,27 @@ mm_impute <-
 
 imputed_aft <- survreg(Surv(imputed_time_to_1500) ~ drug, data = mm_impute)
 summary(imputed_aft)
-saveRDS(imputed_aft, "derived_data/colon_imputed_aft.rds")
+
+pdf("figures/imputed_aft_forest.pdf")
+plot_model(imputed_aft, rm.terms = "Log(scale)", title = "", sort.est = T, axis.title = c("Acceleration Factors", ""))
+dev.off()
+
 ggsurvplot(
   survfit(Surv(imputed_time_to_1500 * 30) ~ drug, data = mm_impute),
   linetype = c("dashed", rep("solid", 7)),
   censor.shape = "|"
 )
+
+## Compare the two models on the same plot
+pdf("figures/aft_models_overlaid_forest.pdf")
+plot_models(
+  aft,
+  imputed_aft,
+  rm.terms = "Log(scale)",
+  title = "",
+  axis.title = c("Acceleration Factors", ""),
+  m.labels = c("Naive", "Imputed"),
+  legend.title = "Models",
+  axis.lim = c(0.5, 3)
+)
+dev.off()

### 2022-04-21 src/R/mixed.R
removed other models and added forest plot figure

@@ -1,6 +1,7 @@
 ## Analysis of colon PDX growth using mixed models
 library(tidyverse)
 library(lme4)
+library(glmmTMB)
 growth <- read.csv("derived_data/pdx_colon_clean.csv") %>%
   mutate(drug = relevel(factor(AgentName), ref = "Control"),
          months_since_first = days_since_first/30)
@@ -11,23 +12,6 @@ fit0 <- glm(
   data = growth,
   family = gaussian(link = "log"))
 summary(fit0)
-
-# ## Random Intercept by mouse
-# fit1 <- glmer(
-#   TUMOR_WT ~ OBS_DAY + OBS_DAY:drug + (1|ID),
-#   data = growth,
-#   family = gaussian(link = "log")
-# )
-# summary(fit1)
-#
-# ## Random slope and intercept by mouse
-# fit2 <- glmer(
-#   TUMOR_WT ~ OBS_DAY + OBS_DAY:drug + (1 + OBS_DAY|ID),
-#   data = growth,
-#   family = gaussian(link = "log"),
-#   glmerControl(optimizer = "bobyqa")
-# )
-# summary(fit2)
 
 ## Random slope and intercept by mouse -- by month
 fitM <- glmer(
@@ -47,35 +31,13 @@ fit_first <- glmer(
 summary(fit_first)
 saveRDS(fit_first, "derived_data/colon_month_first_glmer.rds")
 
-## Random slope and intercept by mouse nested in experiment -- by month
-fitNest <- glmer(
-  TUMOR_WT ~ month + month:drug + (1 + month|ExpNameCSV/ID),
-  data = growth,
-  family = gaussian(link = "log")
+
+pdf("figures/mixed_forest.pdf")
+plot_model(
+  fit_first,
+  transform = "exp",
+  rm.terms = "months_since_first",
+  title = "",
+  axis.title = c("Relative Growth Rate", "")
 )
-summary(fitNest)
-
-## Exploring some different model fitting methods
-library(nlme)
-library(MASS)
-## log-linear model fit with PQL
-fit_day_pql <-
-  MASS::glmmPQL(
-    TUMOR_WT ~ OBS_DAY + OBS_DAY:drug,
-    random = ~ OBS_DAY |
-      ID,
-    family = gaussian(link = "log"),
-    data = growth
-  )
-summary(fit_day_pql)
-
-fit_month_pql <-
-  MASS::glmmPQL(
[…]

### 2022-04-21 src/R/plot_effects.R
dont need effects anymore

@@ -1,13 +1,0 @@
-## Forest plots of effects
-library(tidyverse)
-library(sjPlot)
-
-
-aft <- readRDS("derived_data/aft_model.rds")
-plot_model(aft, title = "AFT Model Acceleration Factors",)
-
-mm <- readRDS("derived_data/colon_month_first_glmer.rds")
-plot_model(mm, transform = "exp")
-
-aft_impute <- readRDS("derived_data/colon_imputed_aft.rds")
-plot_model(aft_impute)

### 2022-04-21 src/R/plot_growth.R
plot pdf and small label changes

@@ -23,30 +23,38 @@ ggplot(vol, aes(x = days_since_first, y = log(TUMOR_WT), color = ExpNameCSV)) +
   labs(x = "Observation Day", y = expression(paste("Log Tumor Volume (m", m^3, ")")))
 
 ## Readjusted time origin facet by drug
+pdf("figures/growth_rescaled_days_drug_facet.pdf")
 ggplot(vol, aes(x = days_since_first, y = log(TUMOR_WT))) +
   geom_line(alpha = 0.5, aes(group = ID)) +
   geom_smooth(method = 'lm', formula = y~x, se = F) +
   facet_wrap(~ AgentName) +
   guides(color = "none") +
   labs(x = "Observation Day", y = expression(paste("Log Tumor Volume (m", m^3, ")")))
+dev.off()
 
 ## Readjust time origin and slopes for each drug
+pdf("figures/growth_rescaled_days.pdf")
 ggplot(vol, aes(x = days_since_first, y = log(TUMOR_WT))) +
   geom_line(alpha = 0.2, aes(group = ID)) +
   geom_smooth(method = 'lm', formula = y~x, se = F, aes(color = AgentName)) +
-  labs(x = "Observation Day", y = expression(paste("Log Tumor Volume (m", m^3, ")")))
+  labs(x = "Days Since Staging", y = expression(paste("Log Tumor Volume (m", m^3, ")")))
+dev.off()
 
 ## Maximum volume distribution
+pdf("figures/max_vol_hist.pdf")
 ggplot(vol, aes(x = maxVol)) +
   geom_histogram(bins = 20, color = "black", fill = "lightblue") +
   geom_vline(xintercept = quantile(vol$maxVol, 0.25), linetype = "dashed") +
   labs(x = expression(paste("Tumor Volume (m", m^3, ")")),
-       y = "Frequency",
-       title = "Maximum Achieved Tumor Volume")
+       y = "Frequency")
+       # title = "Maximum Achieved Tumor Volume")
+dev.off()
 
 ## Distribution of number of follow ups
+pdf("figures/total_obs_hist.pdf")
 ggplot(vol, aes(x = total_obs)) +
   geom_histogram(bins = 15, color = "black", fill = "lightblue") +
   labs(x = "Number of Observations",
-       y = "Frequency",
-       title = "Observations per Mouse")
+       y = "Frequency")
+       # title = "Observations per Mouse")
+dev.off()

### 2022-04-21 paper_Kuhlers.Rmd
more writing

@@
 ---
-title: "Analysis of Patient Derived Xenograft Growth Curves"
+title: "Model Based Analysis of Patient Derived Xenograft Growth Curves"
 author: "Peyton Kuhlers"
-date: "4/6/2022"
-output: pdf_document
+date: "`r Sys.Date()`"
+output:
+  pdf_document:
+    extra_dependencies: ["setspace"]
 ---
@@
 ```
+
+\doublespacing
 
@@
 
-Mice were randomized to treatment groups and did not begin observation
-until reaching a tumor volume of 150-200mm^3 (Evrard et al).
-
-### Data Processing
+### Data Acquisition and Processing
 
@@
 queried for all available colon adenocarcinoma.
+All data were contributed from the NCI Patient-Derived Models Repository (PDMR).
+Mice were staged to a tumor volume of 200mm^3^ (Evrard et al, NCI SOP, Alley et al)
+before being randomized to a treatment group.
+Tumors were measured by caliper 1-3 times per week,
+potentially more depending on the health of animal (Evrard et al, NCI SOP).
+The exact study endpoints for this data were not documented,
+however Evrard et al performed a similar study with data from the PDMR,
+where the study endpoints were "Animal health, Max. tumor size, 4000mm^3^,
+or 300 days, if max. TV not reached" (See Table 1 of Evrard et al).
+Additionally, the PDMR standard operating procedures
+recommend harvesting tumors at 1000-2000mg.
+The distribution of maximum tumor volumes in this dataset reflects this.
 Observation days were rescaled to the number of days the sample
@@
 Fixed effects included were time and the time-drug interaction.
-The drug main effect was not included to model equal mean weight
+The drug main effect was not included to model equal mean volume
 at baseline amongst all treatment groups.
+A random intercept and random time slope were included for each mouse.
+
+### "Naive" Accelerated Failure Time Model
+
+In addition to analyzing tumor volumes directly, time-to-event was
+also analyzed. Since only 2 mice achieved 4000mm^3^,
+the volume endpoint was readjusted 1500mm^3^,
+which was the 25^th^ percentile of volumes in this study.
+A censoring indicator was constructed for whether or not
+1500mm^3^ was achieved.
+Time was taken as the day with the closest measurement to 1500mm^3^.
+With these data, a Weibull accerated failure time (AFT) model was fit.
+
+### "Imputed" Accelerated Failure Time Model
+
+Since some mice exceed 1500mm^3^ (potentially by a substantial amount)
+a "model imputed" time-to-event analysis was also considered.
+If, for example, a mouse achieved 1500mm^3^ on day 30,
+but was not measured until day 40 at 1700mm^3^,
+we can consider time as "measured with error" and will
+over-estimate the treatment effect.
+Conversely, if we only measure 1000mm^3^ on day 20,
+we will be under-estimating the treatment effect.
+To address this, the above mixed effects model was solved
+for time, using the mouse-specific random effects
+to obtain a "model-imputed" time to 1500mm^3^ for each mouse.
+These model imputed estimates were then used to fit
+a Weibull AFT model. For this analysis it was assumed all mice reached
+1500mm^3^ (i.e. no censoring).
 
@@
 \begin{centering}
-\includegraphics[width = 6in, height = 8in]{figures/growth_plots.pdf}
+\includegraphics[width = 6in, height = 8in]{figures/growth_rescaled_days.pdf}
+\end{centering}
+
+\begin{centering}
+\includegraphics[width = 6in, height = 8in]{figures/mixed_forest.pdf}
+\end{centering}
+
+\begin{centering}
+\includegraphics[width = 6in, height = 8in]{figures/aft_models_overlaid_forest.pdf}
 \end{centering}

### 2022-04-26 src/R/aft.R
modified plots

@@ -16,7 +16,15 @@ aft <- survreg(Surv(days_to_1500, reached_1500) ~ drug, data = growth %>% filter
 summary(aft)
 
 pdf("figures/naive_aft_forest.pdf")
-plot_model(aft, rm.terms = "Log(scale)", title = "", sort.est = T, axis.title = c("Acceleration Factors", ""))
+plot_model(
+  aft,
+  rm.terms = "Log(scale)",
+  title = "",
+  axis.labels = rev(levels(growth$drug)[-1]),
+  vline.color = "black",
+  sort.est = T,
+  axis.title = c("Acceleration Factors", "")
+)
 dev.off()
 
 ## Kaplan Meier plot to complement the AFT
@@ -58,7 +66,15 @@ imputed_aft <- survreg(Surv(imputed_time_to_1500) ~ drug, data = mm_impute)
 summary(imputed_aft)
 
 pdf("figures/imputed_aft_forest.pdf")
-plot_model(imputed_aft, rm.terms = "Log(scale)", title = "", sort.est = T, axis.title = c("Acceleration Factors", ""))
+plot_model(
+  imputed_aft,
+  rm.terms = "Log(scale)",
+  title = "",
+  axis.labels = rev(levels(growth$drug)[-1]),
+  vline.color = "black",
+  sort.est = T,
+  axis.title = c("Acceleration Factors", "")
+)
 dev.off()
 
 ggsurvplot(
@@ -75,6 +91,8 @@ plot_models(
   rm.terms = "Log(scale)",
   title = "",
   axis.title = c("Acceleration Factors", ""),
+  axis.labels = rev(levels(growth$drug)[-1]),
+  vline.color = "black",
   m.labels = c("Naive", "Imputed"),
   legend.title = "Models",
   axis.lim = c(0.5, 3)

### 2022-04-26 src/R/mixed.R

@@ -1,14 +1,16 @@
 ## Analysis of colon PDX growth using mixed models
 library(tidyverse)
 library(lme4)
+library(sjPlot)
 library(glmmTMB)
+theme_set(theme_classic())
 growth <- read.csv("derived_data/pdx_colon_clean.csv") %>%
   mutate(drug = relevel(factor(AgentName), ref = "Control"),
          months_since_first = days_since_first/30)
 
 ## OLS estimates
 fit0 <- glm(
-  TUMOR_WT ~ month + month:drug,
+  TUMOR_WT ~ months_since_first + months_since_first:drug,
   data = growth,
   family = gaussian(link = "log"))
 summary(fit0)
@@ -31,13 +33,14 @@ fit_first <- glmer(
 summary(fit_first)
 saveRDS(fit_first, "derived_data/colon_month_first_glmer.rds")
 
-
 pdf("figures/mixed_forest.pdf")
 plot_model(
   fit_first,
   transform = "exp",
   rm.terms = "months_since_first",
   title = "",
-  axis.title = c("Relative Growth Rate", "")
+  axis.title = c("Relative Growth Rate", ""),
+  axis.labels = rev(levels(growth$drug)[-1]),
+  vline.color = "black"
 )
 dev.off()

### 2022-04-26 src/R/plot_growth.R
time distribution plot

@@ -50,6 +50,14 @@ ggplot(vol, aes(x = maxVol)) +
        # title = "Maximum Achieved Tumor Volume")
 dev.off()
 
+## Distribution of time to 1500
+pdf("figures/time_to_1500_hist.pdf")
+ggplot(vol, aes(x = days_to_1500)) +
+  geom_histogram(bins = 20, color = "black", fill = "lightblue") +
+  labs(x = expression(paste("Days to 1500m", m^3)),
+       y = "Frequency")
+dev.off()
+
 ## Distribution of number of follow ups
 pdf("figures/total_obs_hist.pdf")
 ggplot(vol, aes(x = total_obs)) +

### 2022-04-26 presentation_Kuhlers.Rmd
Final presentation

@@
 date: "4/25/2022"
-output: beamer_presentation
+output:
+  beamer_presentation:
+    theme: "AnnArbor"
 ---
@@
 - Given PDXs are meant for high-throughput discovery,
-robust and **efficient** methods are necessary
-  + Inefficient statistical methods waste resources
+we want to use methods that use all available data and maximize the
+chance at finding an effect, while minimizing resource use
 
@@
 
-- Growth data for colon cancer includes ~2700 volume measurements,
-7 drugs (and 1 control), and 263 mice originating from 7 unique patients.
+- ~2600 volume measurements, 7 drugs (and 1 control),
+and 263 mice originating from 7 unique colon cancer patients.
 
-- Mice are randomized to treatments and 'staged' to 150-200mm^3.
+- Generated by the NCI Patient Derived Models Repository (PDMR)
 
-- 
-\centering
-\includegraphics[width = 9cm, height = 5.5cm]{figures/growth_plots.pdf}
-
-## Analysis
-
-- Mixed effects model of tumor growth
+- Mice 'staged' to 150-200mm^3^ then randomized to treatment
 
 \centering
-$log(E[Y_{i}|b_i]) = \beta_0 + b_{i1} + \beta_1(time_{i}) + b_{i2}(time) + \gamma_i (agent_i*time)$
+\includegraphics[width = 7.5cm, height = 5cm]{figures/growth_rescaled_days.pdf}
 
-## Result Reporting
+## Proposed Models
 
-- Profile Plots (shown above)
+- Mixed effects model of growth over time
 
-- Forest plot of interaction coefficients
+- "Naive" accelerated failure time model
 
-- Table of specific pairwise contrasts between different therapies
+- "Imputed" accelerated failure time model
 
-- Variance of random effects
+## Mixed Effects Model
 
-- As a secondary analysis, test methods from Evrard et al 
-at e.g. day 75 and compare to model contrasts at day 75.
+- Is the growth rate significantly different between treatment and control?
+
+\begin{center}
+$log(E[Y_i|b_i]) = \beta_0 + b_{i1} + \beta_1(time_{i}) + b_{i2}(time_i) + \gamma_i (agent_i*time_i)$
+\end{center}
+
+- $Y_i$ is the volume for the i^th^ mouse - assumed Normal
+- Interested in inference on $\gamma_i$
+
+## Mixed Effects Model - Results
+
+- Drug-time interactions -- estimate the change in treatment vs control over time
+- Gemcitabine and Irinotecan weakly reduce growth rates relative to control
+
+\centering
+\includegraphics[width = 8cm, height = 5.5cm]{figures/mixed_forest.pdf}
+
+## Accelerated Failure Time Models
+
+- Do treatments extend the time to some volume endpoint?
+  + PDMR uses 4000mm^3^ as endpoint (really big) - in practice 1000-2000mm^3^
+
+\begin{center}
+$log(T_i) = \beta_0 + \beta_1(agent_i) + \sigma\epsilon$
+\end{center}
+
+- $T_i$ is the (possibly censored) time to 1500mm^3^ for the i^th^ mouse, assumed Weibull
+- Interested in inference on $\beta_1$
+
+## Accelerated Failure Time Models - Cont.
+
+- Time is measured "with error"
+  + "Naive" model - use the time with volume closest to 1500
+  + "Imputed" model - impute time to 1500 by solving the mixed effects model for time
+
+## Accelerated Failure Time Models - Results
+
+- Acceleration factors - Ratio of the mean time to 1500 in treatment vs control
+- Imputed time does not seem to dramatically change estimates
+
+\centering
+\includegraphics[width = 8cm, height = 5.5cm]{figures/aft_models_overlaid_forest.pdf}
+
+## Conclusions
+
+- Gemcitibine and irinotecan had weak effects on growth rates relative to control
+- All drugs (except Vemurafenib) extended the mean time to 1500mm^3^ by ~ 1.3 times
+  + Equal mean at baseline assumption?
+- Imputing time with a mixed model does not dramatically change acceleration estimates
+
+## Limitations
+
+- Exact details of the experiments unknown
+  + Gleaned information from NCI SOPs and other publications
+  + Could be these strains are all resistant to treatment
+
+- More complex relationships were likely missed using a log-linear model
+  + Some drugs may have more complex (nonlinear) effects on growth
+
+- Addressed the naive AFT model with an imputed AFT model
+  + Standard errors of imputed AFT are likely incorrect - bootstrap
+  + Informative censoring?
+
+- The number of times a mouse is measured is related to its well-being
+  + Could produce biased estimates
 

### 2022-04-26 driver.sh
driver update

@@ -3,3 +3,8 @@
 Rscript src/R/clean.R
 
 Rscript src/R/plot_growth.R
+Rscript src/R/aft.R
+Rscript src/R/mixed.R
+
+Rscript -e "rmarkdown::render('presentation_Kuhlers.Rmd')"
+Rscript -e "rmarkdown::render('paper_Kuhlers.Rmd')"

### 2022-04-27 paper_Kuhlers.Rmd
more edits

@@
 are "two dimensional" and all three dimensional interactions are
-lost with cell lines (**REFS**).
-PDX models preserve the 3D nature of human tumors and do not require
+lost with cell lines (Tentler, Pompili)
+PDX models preserve the spatial nature of human tumors and do not require
 transformation, thereby also preserving the genomic characteristics of the
-donor tissue (**REFS**).
+donor tissue (Tentler)
+
+The PDX Development and Trial Centers Research Network (PDXNet) is an
+NCI sponsored consortium aimed at using PDX models to investigate
+cancer therapies and reduce the time therapies move from pre-clinical
+to clinical stages (PDXnet Website). Much of the work from the PDXNet
+has focused on genomic characterization,
 

### 2022-04-30 paper_Kuhlers.Rmd
more writing

--- 2022-04-21 paper_Kuhlers.Rmd
@@
   pdf_document:
-    extra_dependencies: ["setspace"]
+    extra_dependencies: ["setspace", "float"]
 ---
@@
 
+Abstract
+========
+
+Patient derived xenografts (PDX) are preclinical cancer models frequently used
+as platforms to investigate novel therapies and identify biomarkers.
+Tumor growth over time are often recorded during PDX experiments, but
+this data is typically analyzed with relatively simple methods such as percent change.
+This analysis investigates 2 model based methods for the analysis of PDX growth
+data. A mixed effects model was used to investigate the effect of 7 different
+drugs on tumor growth over time. Gemcitabine and irinotecan were found to
+reduce tumor growth relative to control with relative growth rates of 0.60 (95% CI: 0.43, 0.85) 
+and 0.71 (95% CI: 0.51, 0.99), respectively. An accelerated failure time
+model (AFT) was used to investigate the mean time to 1500mm^3^. Except for vemurafenib,
+mean times to 1500mm^3^ were about 1.3 times longer in drug treated mice
+compared to control mice. Carboplatin and erlotinib extended time to 1500mm^3^
+the longest relative to control with acceleration factors of 1.40 (95% CI: 1.34, 1.46)
+and 1.39 (95% CI: 1.32, 1.46), respectively. A time-imputed AFT model was used to investigate
+the effect of error in the measurement of the time to 1500mm^3^. Acceleration
+factors remained largely unchanged between naive and imputed AFT models,
+indicating time to 1500mm^3^ was measured relatively accurately.
+
 Introduction
@@
 
-Patient Derived Xenografts (PDX) are *in vivo* cancer models
+Patient derived xenografts (PDX) are *in vivo* cancer models
 in which patient tumors are removed and engrafted into immuno-deficient mice.
-Once established, PDXs can be used as platforms for drug discovery or
-investigation of *in vivo* drug mechanisms (**REFS**).
-
+Once established, PDXs can be used as platforms for drug and biomarker discovery or
+for investigation of drug mechanisms *in vivo* (**REFS**).
 PDX models have significant advantages over alternative models.
@@
 are "two dimensional" and all three dimensional interactions are
-lost with cell lines (**REFS**).
-PDX models preserve the 3D nature of human tumors and do not require
+lost with cell lines (REF Tentler, Pompili)
+PDX models preserve the spatial nature of human tumors and do not require
 transformation, thereby also preserving the genomic characteristics of the
-donor tissue (**REFS**).
+donor tissue (REF Tentler)
+
+The PDX Development and Trial Centers Research Network (PDXNet) is an
+NCI sponsored consortium aimed at using PDX models to investigate
+cancer therapies and reduce the time therapies move from pre-clinical
+to clinical stages (REF PDXnet Website). Much of the work from the PDXNet
+has focused on genomic characterization and studies identifying
+vulnerabilities and resistances to treatment (REFS).
+While these studies often assess tumor growth over time,
+the methods used to test differences are often limited
+to t-tests and ANOVA (REFS). Evrard et al surveyed a variety of methods
+for analyzing tumor volumes including percent change, area under the growth
+curve, and RECIST (a criteria for evaluating treatment response).
+Aside from area under the curve, these metrics do not leverage all
+of the available data. Since tumors are generally measured 2-3 times per week,
+there is a wealth of data for assessing growth trajectories in between
+baseline and the final measurement (REF Evrard, NCI?).
+Pearson et al and Oberg et al are two examples of model based approaches
+for analyzing PDX tumor growth. Pearson et al used mixed effects
+models to analyze growth data over multiple tumor passages and Oberg et al
+detailed strategies for using mixed effects models to evaluate
+treatments in PDXs.
+The aim of this analysis is to explore a mixed model based approach for analyzing
+PDX tumor volumes, similar to Pearson et al and Oberg et al.
+While mixed models impose more assumptions on the structure of data than
+e.g. a t-test, they leverage all of the available data,
+can handle missing and uneven follow-up times, and have flexibility
+to model a variety of growth curve shapes.
+In addition to a mixed effects model analysis, an accelerated failure time (AFT)
+model approach is also explored.
 
@@
 Additionally, the PDMR standard operating procedures
-recommend harvesting tumors at 1000-2000mg.
+recommend harvesting tumors at 1000-2000mg (REF PDMR SOP).
 The distribution of maximum tumor volumes in this dataset reflects this.
@@
 days since day 30. Days were then rescaled to months.
+All data processing and analysis were conducted using R/v4.1.0 (REF R core).
 
@@
 at baseline amongst all treatment groups.
+Volumes were assumed to be normally distributed and log-linearly
+related to the covariates.
 A random intercept and random time slope were included for each mouse.
+Models were fit using the R package `lme4`/v1.1.27.
+All confidence intervals are 95% Wald-type intervals.
 
@@
 the volume endpoint was readjusted 1500mm^3^,
-which was the 25^th^ percentile of volumes in this study.
+which was approximately the 25^th^ percentile of maximum volumes in this study.
 A censoring indicator was constructed for whether or not
@@
 Time was taken as the day with the closest measurement to 1500mm^3^.
-With these data, a Weibull accerated failure time (AFT) model was fit.
+With these data, a Weibull accelerated failure time model was fit
+with the R package `survival`/v3.2.13 (REF R package)
 
@@
 we will be under-estimating the treatment effect.
-To address this, the above mixed effects model was solved
+To address this, the above mixed effects model was set to log(1500) and solved
 for time, using the mouse-specific random effects
@@
 
-\begin{centering}
+This dataset consisted of 2610 volume measurements from 263 mice originating
+from 7 colon adenocarcinoma patients. There was a median of 11 volume
+measurements (IQR: 8-15) taken per mouse.
+Mice were measured for a median of 36 days (IQR: 27-50).
+The median final volume for all mice was 1691mm^3^ (IQR: 1542-1974),
+which is in line with recommendations from the PDMR on harvesting volumes (REF NCI).
+In general, log scaled tumor volumes grew linearly with time,
+however, there were some mice with non-linear growth curves on the log scale (Fig. 1).
+As seen in Fig. 1, simple linear regression lines seem to indicate
+that there is no particularly effective treatment, as most of the slopes
+are approximately parallel to the control group (flat or negative slopes
+are expected for highly active treatments). Strangely, vemurafenib treated
+tumors appear to be growing faster than control.
+
+\begin{center}
+\begin{figure}[H]
 \includegraphics[width = 6in, height = 8in]{figures/growth_rescaled_days.pdf}
-\end{centering}
+\caption{Log tumor volumes over time.
+Black lines indicate each mouses' growth curve.
+Colored lines are simple linear regressions of log-volume versus time,
+separated by each treatment group}
+\end{figure}
+\end{center}
 
-\begin{centering}
+The aim of the mixed effects model analysis was to model the trajectory
+of the growth curve and determine if treatments slow tumor growth.
+Therefore the statistic of interest is the slope,
+which in the mixed model represents the change in the ratio of the
+mean tumor volume in treatment versus control per month.
+Only gemcitabine and irinotecan were found to significantly reduce
+tumor growth rates compared to control. The relative growth rates
+were 0.60 (95% CI: 0.43, 0.85) and 0.71 (95% Wald CI: 0.51, 0.99) which correspond to a
+40% and 29% reduction in the growth rate relative to control, respectively (Fig. 3).
+
+\begin{center}
+\begin{figure}[H]
 \includegraphics[width = 6in, height = 8in]{figures/mixed_forest.pdf}
-\end{centering}
+\caption{Forest plot of exponentiated mixed effects model interaction coefficients.
+The coefficients represent the change in the ratio of mean volume in treatment versus control
+over time. Red points indicate the growth rate is reduced in the treatment group,
+while blue points indicate the growth rate is increased in the treatment group.
+Horizontal bars indicate 95\% confidence intervals.}
+\end{figure}
+\end{center}
 
-\begin{centering}
-\includegraphics[width = 6in, height = 8in]{figures/aft_models_overlaid_forest.pdf}
-\end{centering}
+The aim of the AFT analysis was to determine if treatment could extend the
+time to a specific volume endpoint. This analysis can be seen as complementary
+to analyzing area under the curve. While complex non-linear growth curves may
+be difficult to adequately describe with a regression model (see Fig. 1),
+the AFT model summarizes the curve as simply the time to the endpoint.
+In contrast with the mixed model, all treatments -- aside from vemurafenib --
+appear effective and extend the mean time to 1500mm^3^
+by about 1.3 times relative to control.
+Carboplatin and erlotinib extended the mean time to 1500mm^3^ by 1.40 (95% CI: 1.34, 1.46)
+and 1.39 (95% CI: 1.32, 1.46) times, respectively.
+As aforementioned, the naive AFT model uses the time with volume closest
+to 1500mm^3^, which can over or under estimate the treatment effect.
+Therefore an imputed model was tested to determine if the naive model was
+sensitive to the exact time to 1500mm^3^. Surprisingly, aside from 5-FU imputation
+had a very minor effect on estimates of the acceleration factors (Fig. 3)
+
+\begin{center}
+\begin{figure}[H]
+\includegraphics[width = 6in, height = 5in]{figures/aft_models_overlaid_forest.pdf}
+\caption{Forest plot of AFT acceleration factors.
+These represent the ratio of the mean time to 1500mm$^3$ in treatment versus control.
+Red points indicate that time to 1500mm$^3$ is extended relative to control,
+while blue points indicate the time is shortened.}
+\end{figure}
+\end{center}
 
@@
 ===========
+
+The aim of this analysis was to use modelling based approaches for analyzing the growth
+of PDX tumors treated with various chemotherapies.
+Two models were utilized: a generalized mixed model and an accelerated failure time model.
+Results from the mixed model show that all the drugs--aside from vemurafenib--
+reduced the growth rate relative to control, however,
+the reductions range from modest to essentially null (Fig. 2).
+Irinotecan and 5-FU are two frequently used drugs approved by the FDA for the treatment of colon cancer.
+Each have been shown to improve overall survival (REF labianca, fujita, nci)
+and had comparable effect sizes in this analysis.
+Gemcitabine is not FDA approved for colon cancer, but further study
+may be warranted as it was found to modestly reduce tumor growth rates in this analysis.
+Vemurafenib is an outlier in this dataset, being the only drug that showed
+increased tumor growth relative to control.
+Vemurafenib has been shown to improve progression free survival in patients
+with a *BRAF*^V600E^ mutation (REF kopetz), but is not FDA approved for
+colon cancer. It is possible that the group of
+vemurafenib treated tumors were particularly fast growing and clearly resistant
+to treatment.
+
+AFT models were also explored in this analysis. As aforementioned,
+AFT models could be a complementary approach to area under the growth curve (AUC).
+Both AUC and AFT analyses summarize the growth curve to a single number -- AUC or time to the endpoint
+which can be useful with the growth curves have complex shapes.
+For example, as seen in Fig. 1, some mice experience a period of tumor regression
+followed by rapid tumor growth. Modeling this relationship is difficult
+and may quickly become ad-hoc and produce overfitted models (especially
+if the relationship is only observed in a handful of mice). An AUC or AFT analysis
+can relatively easily handle this relationship since these mice will simply have
+larger AUC and longer event times. The advantage of the AFT model is the results
+are easily interpretable compared to AUC and can incorporate censoring for mice
+that do not reach the volume endpoint.
+
+The AFT models in this analysis largely reflected the mixed model results.
+All drugs aside from vemurafenib extended the time to 1500mm^3^, with
+carboplatin and erlotinib having the largest effect size.
+While these results corresponds to the observed growth rates from the mixed model,
+all of the acceleration factors were significant, while only a few of
+the relative growth rates were. 
+The discrepancy in effective drugs found by the two models could potentially be explained
+by differences in baseline tumor volume. If two groups have the same slope
+but different intercepts, the group with the higher intercept will reach
+the endpoint first simply by being heavier at baseline, not because the treatment was effective.
+While the experimental design and protocol justify the assumption,
+its possible in practice inefficiencies in staging mice result in group level
+differences at baseline. Additionally, the linear relationship may be poor approximations
+of the true growth trajectory causing true differences in growth rates to be masked.
+The discrepancy could also be simply
+because small numerical differences in the growth rates (while statistically insignificant)
+result in the overall time to 1500mm^3^ to be different.
+
+Time to 1500mm^3^ is an artificial endpoint used for exploratory purposes in this analysis
+and future experiments could set more meaningful volume endpoints that
+may dramatically change these results.
+
+Interestingly, imputing the time to 1500mm^3^ using estimates from the mixed model do not
+dramatically change acceleration factor. The interquartile range of the difference
+between imputed and naive times to 1500 was -0.15 to 0.01, indicating
+the time to 1500mm^3^ was measured fairly accurately, resulting in stable
+estimates between the two models.
+Whether these results correspond to any outcomes experienced by the donor patients is unknown,
+and future work could incorporate patient level outcomes to investigate
+the consistency between PDX and patient outcomes.
+
+### Limitations
+
+This study has several limitations.
+The exact details of the experiment procedure for this data was not detailed
+with the dataset. Other sources such as PDMR standard operating procedures (SOP)
+and Evrard et al had to be used to determine the protocol and potential endpoints.
+Data such as the treatment sensitivity of the strain or the dosing protocol
+could explain why relatively weak effect sizes were observed.
+
+This analysis only incorporated linear terms into the mixed effects model,
+however, it is likely that there are more complex relationships
+that are caused by the drug mechanism or dosing regime (Fig. 1).
+Qualitative assessment of the model revealed that the marginal predictions are quite
+poor (for example, the marginal model estimated control mice will be 75,000mm^3^ at 3 months,
+which is impossible).
+Future work could explore ways to more flexibly model the growth trajectory
+or explore ways to weight mice with more observations over a longer duration
+to get better estimates of the growth rate.
+Additional work could explore censoring-like indicators to account for mice
+having short observations times due to experimental or ethical constraints,
+rather than simply being missing. (REF Tan)
+
+While this analysis addressed the issue of "error" in assessment of the time
+to 1500mm^3^ by using a mixed model imputed time, this approached introduced another
+layer of complexity. In this analysis standard errors from the imputed analysis
+were taken as-is. However, the standard errors are likely larger since we are
+imputing a model derived estimate. Future work could simply use the bootstrap to
+produce more honest standard error estimates.
+
+Lastly, Evrard et al report "Animal Health" as one of the primary study endpoints
+for PDMR data. According to PDMR SOPs, mice may be weighed more frequently
+if "clinical signs are noted" (REF PDMR SOP). Therefore,
+the number of measurements on the mouse may be related to the health
+and overall outcome of the mouse, which may potentially bias estimates
+from the mixed model. Future work could assess how informative observation numbers
+impact estimates of drug efficacy.
+
+References
+==========
+
+1) Tentler JJ, Tan AC, Weekes CD, et al.
+Patient-derived tumour xenografts as models for oncology drug development.
+Nat Rev Clin Oncol.
+2012;9(6):338-350.
+Published 2012 Apr 17.
+doi:10.1038/nrclinonc.2012.61
+The NCI Patient-Derived Models Repository (PDMR), NCI-Frederick, Frederick National Laboratory for Cancer Research, Frederick, MD. https://pdmr.cancer.gov/. 
+Drugs approved for colon and rectal cancer. National Cancer Institute. https://www.cancer.gov/about-cancer/treatment/drugs/colorectal#1. Accessed April 30, 2022. 
+Tan M, Fang HB, Tian GL, Houghton PJ. Repeated-measures models with constrained parameters for incomplete data in tumour xenograft experiments. Stat Med. 2005;24(1):109-119. doi:10.1002/sim.1775
+Fujita K, Kubota Y, Ishida H, Sasaki Y. Irinotecan, a key chemotherapeutic drug for metastatic colorectal cancer. World J Gastroenterol. 2015;21(43):12234-12248. doi:10.3748/wjg.v21.i43.12234
+Labianca R, Beretta GD, Kildani B, et al. Colon cancer. Crit Rev Oncol Hematol. 2010;74(2):106-133. doi:10.1016/j.critrevonc.2010.01.010
+Kopetz S, Guthrie KA, Morris VK, et al.
+Randomized Trial of Irinotecan and Cetuximab With or Without Vemurafenib in BRAF-Mutant Metastatic Colorectal Cancer (SWOG S1406).
+J Clin Oncol.
+2021;39(4):285-294.
+doi:10.1200/JCO.20.01994

### 2022-05-01 paper_Kuhlers.Rmd
more edits and citations

@@
   pdf_document:
-    extra_dependencies: ["setspace", "float"]
+    extra_dependencies: ["setspace", "float", "fancybox", "graphicx"]
+bibliography: citations.bib
+csl: https://raw.githubusercontent.com/citation-style-language/styles/master/american-medical-association.csl
 ---
@@
 Once established, PDXs can be used as platforms for drug and biomarker discovery or
-for investigation of drug mechanisms *in vivo* (**REFS**).
+for investigation of drug mechanisms *in vivo*[@tentler_patient-derived_2012; @pompili_patient-derived_2016].
 PDX models have significant advantages over alternative models.
@@
 are "two dimensional" and all three dimensional interactions are
-lost with cell lines (REF Tentler, Pompili)
+lost with cell lines.
 PDX models preserve the spatial nature of human tumors and do not require
 transformation, thereby also preserving the genomic characteristics of the
-donor tissue (REF Tentler)
+donor tissue[@tentler_patient-derived_2012; @pompili_patient-derived_2016; @daniel_primary_2009].
 
@@
 cancer therapies and reduce the time therapies move from pre-clinical
-to clinical stages (REF PDXnet Website). Much of the work from the PDXNet
+to clinical stages[@noauthor_notitle_nodate]. Much of the work from the PDXNet
 has focused on genomic characterization and studies identifying
-vulnerabilities and resistances to treatment (REFS).
+vulnerabilities and resistances to treatment.
 While these studies often assess tumor growth over time,
 the methods used to test differences are often limited
-to t-tests and ANOVA (REFS). Evrard et al surveyed a variety of methods
+to t-tests and ANOVA. Evrard et al[@evrard_systematic_2020] surveyed a variety of methods
 for analyzing tumor volumes including percent change, area under the growth
@@
 there is a wealth of data for assessing growth trajectories in between
-baseline and the final measurement (REF Evrard, NCI?).
+baseline and the final measurement[@evrard_systematic_2020; @noauthor_patient-derived_nodate].
 Pearson et al and Oberg et al are two examples of model based approaches
@@
 detailed strategies for using mixed effects models to evaluate
-treatments in PDXs.
+treatments in PDXs[@pearson_patient-derived_2016; @oberg_statistical_2021].
 The aim of this analysis is to explore a mixed model based approach for analyzing
@@
 queried for all available colon adenocarcinoma.
-All data were contributed from the NCI Patient-Derived Models Repository (PDMR).
-Mice were staged to a tumor volume of 200mm^3^ (Evrard et al, NCI SOP, Alley et al)
-before being randomized to a treatment group.
+All data were contributed from the NCI Patient-Derived Models Repository (PDMR)[@noauthor_patient-derived_nodate].
+Mice were staged to a tumor volume of ~200mm^3^ before being randomized to a treatment group.
 Tumors were measured by caliper 1-3 times per week,
-potentially more depending on the health of animal (Evrard et al, NCI SOP).
+potentially more depending on the health of animal[@evrard_systematic_2020; @noauthor_patient-derived_nodate; @teicher_human_2004].
 The exact study endpoints for this data were not documented,
@@
 where the study endpoints were "Animal health, Max. tumor size, 4000mm^3^,
-or 300 days, if max. TV not reached" (See Table 1 of Evrard et al).
+or 300 days, if max. TV not reached" (See Table 1 of Evrard et al[@evrard_systematic_2020]).
 Additionally, the PDMR standard operating procedures
-recommend harvesting tumors at 1000-2000mg (REF PDMR SOP).
+recommend harvesting tumors at 1000-2000mg[@noauthor_patient-derived_nodate].
 The distribution of maximum tumor volumes in this dataset reflects this.
@@
 days since day 30. Days were then rescaled to months.
-All data processing and analysis were conducted using R/v4.1.0 (REF R core).
+All data processing and analysis were conducted using R/v4.1.0 ([@r-core]).
 
@@
 A random intercept and random time slope were included for each mouse.
-Models were fit using the R package `lme4`/v1.1.27.
+Models were fit using the R package `lme4`/v1.1.27[@bates].
 All confidence intervals are 95% Wald-type intervals.
@@
 With these data, a Weibull accelerated failure time model was fit
-with the R package `survival`/v3.2.13 (REF R package)
+with the R package `survival`/v3.2.13[@survival-package].
 
@@
 The median final volume for all mice was 1691mm^3^ (IQR: 1542-1974),
-which is in line with recommendations from the PDMR on harvesting volumes (REF NCI).
+which is in line with recommendations from the PDMR on harvesting volumes[@noauthor_patient-derived_nodate].
 In general, log scaled tumor volumes grew linearly with time,
@@
 
+\begin{figure}
 \begin{center}
-\begin{figure}[H]
-\includegraphics[width = 6in, height = 8in]{figures/growth_rescaled_days.pdf}
+\shadowbox{\includegraphics[width = 6in, height = 5in]{figures/growth_rescaled_days.pdf}}
 \caption{Log tumor volumes over time.
@@
 separated by each treatment group}
+\end{center}
 \end{figure}
-\end{center}
 
@@
 of the growth curve and determine if treatments slow tumor growth.
-Therefore the statistic of interest is the slope,
-which in the mixed model represents the change in the ratio of the
-mean tumor volume in treatment versus control per month.
+Therefore, we are interested in the growth rate (slope) of the tumor growth curve.
+The mixed model coefficients represent the growth rates for each treatment,
+relative to the control groups growth. The coefficients can
+also be interpreted as the change in the ratio of the mean volume in the
+treatment group compared to control per month.
+Values less than 1 indicate the treatment reduces the growth rate relative to control.
 Only gemcitabine and irinotecan were found to significantly reduce
 tumor growth rates compared to control. The relative growth rates
-were 0.60 (95% CI: 0.43, 0.85) and 0.71 (95% Wald CI: 0.51, 0.99) which correspond to a
-40% and 29% reduction in the growth rate relative to control, respectively (Fig. 3).
+were 0.60 (95% CI: 0.43, 0.85) and 0.71 (95% CI: 0.51, 0.99) which correspond to a
+40% and 29% reduction in the growth rate relative to control, respectively (Fig. 2).
 
+\begin{figure}
 \begin{center}
-\begin{figure}[H]
-\includegraphics[width = 6in, height = 8in]{figures/mixed_forest.pdf}
+\shadowbox{\includegraphics[width = 6in, height = 5in]{figures/mixed_forest.pdf}}
 \caption{Forest plot of exponentiated mixed effects model interaction coefficients.
@@
 Horizontal bars indicate 95\% confidence intervals.}
+\end{center}
 \end{figure}
-\end{center}
 
@@
 
+\begin{figure}
 \begin{center}
-\begin{figure}[H]
-\includegraphics[width = 6in, height = 5in]{figures/aft_models_overlaid_forest.pdf}
+\shadowbox{\includegraphics[width = 6in, height = 5in]{figures/aft_models_overlaid_forest.pdf}}
 \caption{Forest plot of AFT acceleration factors.
@@
 while blue points indicate the time is shortened.}
+\end{center}
 \end{figure}
-\end{center}
 
@@
 Irinotecan and 5-FU are two frequently used drugs approved by the FDA for the treatment of colon cancer.
-Each have been shown to improve overall survival (REF labianca, fujita, nci)
+Each have been shown to improve overall survival[@labianca_colon_2010; @fujita_irinotecan_2015; @noauthor_drugs_2011].
 and had comparable effect sizes in this analysis.
@@
 Vemurafenib has been shown to improve progression free survival in patients
-with a *BRAF*^V600E^ mutation (REF kopetz), but is not FDA approved for
+with a *BRAF*^V600E^ mutation[@kopetz_randomized_2021], but is not FDA approved for
 colon cancer. It is possible that the group of
@@
 having short observations times due to experimental or ethical constraints,
-rather than simply being missing. (REF Tan)
+rather than simply being missing[@tan_repeated-measures_2005].
 
@@
 for PDMR data. According to PDMR SOPs, mice may be weighed more frequently
-if "clinical signs are noted" (REF PDMR SOP). Therefore,
+if "clinical signs are noted"[@evrard_systematic_2020; @noauthor_patient-derived_nodate]. Therefore,
 the number of measurements on the mouse may be related to the health
@@
 ==========
-
-1) Tentler JJ, Tan AC, Weekes CD, et al.
-Patient-derived tumour xenografts as models for oncology drug development.
-Nat Rev Clin Oncol.
-2012;9(6):338-350.
-Published 2012 Apr 17.
-doi:10.1038/nrclinonc.2012.61
-The NCI Patient-Derived Models Repository (PDMR), NCI-Frederick, Frederick National Laboratory for Cancer Research, Frederick, MD. https://pdmr.cancer.gov/. 
-Drugs approved for colon and rectal cancer. National Cancer Institute. https://www.cancer.gov/about-cancer/treatment/drugs/colorectal#1. Accessed April 30, 2022. 
-Tan M, Fang HB, Tian GL, Houghton PJ. Repeated-measures models with constrained parameters for incomplete data in tumour xenograft experiments. Stat Med. 2005;24(1):109-119. doi:10.1002/sim.1775
-Fujita K, Kubota Y, Ishida H, Sasaki Y. Irinotecan, a key chemotherapeutic drug for metastatic colorectal cancer. World J Gastroenterol. 2015;21(43):12234-12248. doi:10.3748/wjg.v21.i43.12234
-Labianca R, Beretta GD, Kildani B, et al. Colon cancer. Crit Rev Oncol Hematol. 2010;74(2):106-133. doi:10.1016/j.critrevonc.2010.01.010
-Kopetz S, Guthrie KA, Morris VK, et al.
-Randomized Trial of Irinotecan and Cetuximab With or Without Vemurafenib in BRAF-Mutant Metastatic Colorectal Cancer (SWOG S1406).
-J Clin Oncol.
-2021;39(4):285-294.
-doi:10.1200/JCO.20.01994

### 2022-05-01 paper_Kuhlers.Rmd
figure edits

@@
 days since day 30. Days were then rescaled to months.
-All data processing and analysis were conducted using R/v4.1.0 ([@r-core]).
+All data processing and analysis were conducted using R/v4.1.0[@r-core].
 
@@
 
-\begin{figure}
+\begin{figure}[H]
 \begin{center}
-\shadowbox{\includegraphics[width = 6in, height = 5in]{figures/growth_rescaled_days.pdf}}
+\shadowbox{\includegraphics[width = 6in, height = 4.5in]{figures/growth_rescaled_days.pdf}}
 \caption{Log tumor volumes over time.
@@
 
-\begin{figure}
+\begin{figure}[H]
 \begin{center}
-\shadowbox{\includegraphics[width = 6in, height = 5in]{figures/mixed_forest.pdf}}
+\shadowbox{\includegraphics[width = 6in, height = 4.5in]{figures/mixed_forest.pdf}}
 \caption{Forest plot of exponentiated mixed effects model interaction coefficients.
@@
 
-\begin{figure}
+\begin{figure}[H]
 \begin{center}
-\shadowbox{\includegraphics[width = 6in, height = 5in]{figures/aft_models_overlaid_forest.pdf}}
+\shadowbox{\includegraphics[width = 6in, height = 4.5in]{figures/aft_models_overlaid_forest.pdf}}
 \caption{Forest plot of AFT acceleration factors.

### 2022-06-30 src/R/clean.R
some changes

@@ -1,5 +1,6 @@
 library(tidyverse)
 growth <- read.csv("source_data/pdx_colon.csv", row.names = 1) %>%
+  distinct() %>%
   mutate(
     day = OBS_DAY - 7,
     month = OBS_DAY / 30,
@@ -11,10 +12,8 @@ growth <- read.csv("source_data/pdx_colon.csv", row.names = 1) %>%
     maxVol = max(TUMOR_WT),
     days_since_first = OBS_DAY - first,
     reached_1500 = ifelse(maxVol >= 1500, 1, 0),
-    total_obs = length(OBS_DAY),
+    total_obs = n(),
     days_to_1500 = days_since_first[which.min(1500 - TUMOR_WT)]
-  ) %>%
-  ungroup() %>%
-  distinct()
+  )
 
 write.csv(growth, "derived_data/pdx_colon_clean.csv")

### 2022-06-30 src/R/mixed.R

@@ -2,7 +2,6 @@
 library(tidyverse)
 library(lme4)
 library(sjPlot)
-library(glmmTMB)
 theme_set(theme_classic())
 growth <- read.csv("derived_data/pdx_colon_clean.csv") %>%
   mutate(drug = relevel(factor(AgentName), ref = "Control"),
@@ -38,9 +37,73 @@ plot_model(
   fit_first,
   transform = "exp",
   rm.terms = "months_since_first",
-  title = "",
+  title = "", axis.lim = c(0.3, 2),
   axis.title = c("Relative Growth Rate", ""),
   axis.labels = rev(levels(growth$drug)[-1]),
   vline.color = "black"
 )
 dev.off()
+
+####################################################################
+## Some exploration
+####################################################################
+
+# growth$months_square <- growth$months_since_first^2
+# fit_square <- glmer(
+#   TUMOR_WT ~ poly(months_since_first, 2) + poly(months_since_first,2):drug + (1 + poly(months_since_first, 1)||ID),
+#   data = growth,
+#   family = gaussian(link = "log")
+# )
+# summary(fit_square)
+# 
+# growth$log_month <- log(growth$months_since_first + 1)
+# fit_log <- glmer(
+#   TUMOR_WT ~ log_month + log_month:drug + (1 + log_month|ID),
+#   data = growth,
+#   family = gaussian(link = "log")
+# )
+# growth$idk <- growth$months_since_first + 1
+# 
+# fit_idk <- glmer(
+#   TUMOR_WT ~ idk + idk:drug + (1 + idk|ID),
+#   data = growth,mustart = pmax(growth$TUMOR_WT, 1e-3),
+#   family = gaussian(link = "log")
+# )
+# ## More explore
+# test <- growth %>%
+#   filter(TUMOR_WT < 1600)
+# 
+# growth$tumor_g <- growth$TUMOR_WT/1000
+# uh <- glmer(
+#   tumor_g ~ months_since_first + months_since_first:drug + (1 + months_since_first|ID),
+#   data = growth,
+#   family = gaussian(link = "log")
+# )
+# ## lmer?
+# fit_first_lmer <- lmer(
+#   TUMOR_WT ~ months_since_first + months_since_first:drug + (1 + months_since_first|ID),
+#   data = growth
+# )
+# fit_first_full <- glmer(
+#   TUMOR_WT ~ months_since_first + drug + months_since_first:drug + (1 + months_since_first|ID),
+#   data = growth,
[…]

### 2022-06-30 src/R/plot_growth.R

@@ -24,21 +24,31 @@ ggplot(vol, aes(x = days_since_first, y = log(TUMOR_WT), color = ExpNameCSV)) +
 
 ## Readjusted time origin facet by drug
 pdf("figures/growth_rescaled_days_drug_facet.pdf")
-ggplot(vol, aes(x = days_since_first, y = log(TUMOR_WT))) +
+ggplot(vol, aes(x = (days_since_first), y = log(TUMOR_WT))) +
   geom_line(alpha = 0.5, aes(group = ID)) +
+  # geom_smooth(method = 'loess', formula = y~x, se = F) +
   geom_smooth(method = 'lm', formula = y~x, se = F) +
   facet_wrap(~ AgentName) +
   guides(color = "none") +
-  labs(x = "Observation Day", y = expression(paste("Log Tumor Volume (m", m^3, ")")))
+  labs(x = "Days Since Staging", y = expression(paste("Log Tumor Volume (m", m^3, ")")))
 dev.off()
 
 ## Readjust time origin and slopes for each drug
 pdf("figures/growth_rescaled_days.pdf")
-ggplot(vol, aes(x = days_since_first, y = log(TUMOR_WT))) +
+ggplot(vol, aes(x = (days_since_first), y = log(TUMOR_WT))) +
   geom_line(alpha = 0.2, aes(group = ID)) +
   geom_smooth(method = 'lm', formula = y~x, se = F, aes(color = AgentName)) +
+  # geom_smooth(method = 'loess', formula = y~x, se = F) +
   labs(x = "Days Since Staging", y = expression(paste("Log Tumor Volume (m", m^3, ")")))
 dev.off()
+
+## Readjust time origin + mean profile rather than indiv. profile
+v <- vol %>% group_by(days_since_first, drug) %>% summarize(avgwt = mean(log(TUMOR_WT)))
+ggplot(v, aes(x = days_since_first, y = avgwt)) +
+  geom_line() +
+  geom_smooth(method = "lm", se = F) +
+  ggpubr::stat_regline_equation() +
+  facet_wrap(~ drug)
 
 ## Maximum volume distribution
 pdf("figures/max_vol_hist.pdf")

### 2022-06-30 paper_Kuhlers.Rmd
final doc

@@
 ---
-title: "Model Based Analysis of Patient Derived Xenograft Growth Curves"
-author: "Peyton Kuhlers"
-date: "`r Sys.Date()`"
 output:
   pdf_document:
+    latex_engine: xelatex
+    keep_tex: true
     extra_dependencies: ["setspace", "float", "fancybox", "graphicx"]
+mainfont: Times New Roman
+fontsize: 12pt
 bibliography: citations.bib
@@
 
+\newgeometry{top=2in,left=1in,right=1in}
+\begin{titlepage}
+\begin{center}
+
+  MODEL BASED ANALYSIS OF PATIENT DERIVED XENOGRAFT GROWTH CURVES
+  \vspace{1in}
+
+  Peyton Kuhlers
+  \vspace*{1in}
+
+  A master's paper submitted to the faculty at the University of North Carolina at Chapel Hill
+  in partial fulfillment of the requirements for the degree of Master of Science in the
+  Department of Biostatistics in the Gillings School of Global Public Health.
+  \vspace{1in}
+
+  \begin{singlespace}
+  Chapel Hill\\
+
+  2022
+  \end{singlespace}
+  \vspace{1in}
+
+\begin{doublespace}
+  \begin{flushright}
+    \begin{tabular}{l@{}}
+      Approved by:\\Bahjat Qaqish\\Baiming Zou
+    \end{tabular}
+  \end{flushright}
+\end{doublespace}
+
+\end{center}
+\end{titlepage}
+
+\restoregeometry
 \doublespacing
@@
 indicating time to 1500mm^3^ was measured relatively accurately.
+
+\newpage
 
@@
 at baseline amongst all treatment groups.
-Volumes were assumed to be normally distributed and log-linearly
+Volumes were assumed to be normally distributed with the log of the mean linearly
 related to the covariates.
@@
 \shadowbox{\includegraphics[width = 6in, height = 4.5in]{figures/growth_rescaled_days.pdf}}
+\begin{singlespace}
 \caption{Log tumor volumes over time.
@@
 separated by each treatment group}
+\end{singlespace}
 \end{center}
@@
 The mixed model coefficients represent the growth rates for each treatment,
-relative to the control groups growth. The coefficients can
+relative to the control group's growth. The coefficients can
 also be interpreted as the change in the ratio of the mean volume in the
@@
 \shadowbox{\includegraphics[width = 6in, height = 4.5in]{figures/mixed_forest.pdf}}
+\begin{singlespace}
 \caption{Forest plot of exponentiated mixed effects model interaction coefficients.
@@
 Horizontal bars indicate 95\% confidence intervals.}
+\end{singlespace}
 \end{center}
@@
 \shadowbox{\includegraphics[width = 6in, height = 4.5in]{figures/aft_models_overlaid_forest.pdf}}
+\begin{singlespace}
 \caption{Forest plot of AFT acceleration factors.
 These represent the ratio of the mean time to 1500mm$^3$ in treatment versus control.
-Red points indicate that time to 1500mm$^3$ is extended relative to control,
-while blue points indicate the time is shortened.}
+Blue points indicate estimates from the naive model
+and red points indicate estimates from the imputed model.}
+\end{singlespace}
 \end{center}
@@
 Both AUC and AFT analyses summarize the growth curve to a single number -- AUC or time to the endpoint
-which can be useful with the growth curves have complex shapes.
+which can be useful when growth curves have complex shapes.
 For example, as seen in Fig. 1, some mice experience a period of tumor regression
@@
 The discrepancy in effective drugs found by the two models could potentially be explained
-by differences in baseline tumor volume. If two groups have the same slope
-but different intercepts, the group with the higher intercept will reach
+by differences in baseline tumor volume. If two groups have the same growth rate (slope)
+but different baseline volumes (intercepts), the group with the higher intercept will reach
 the endpoint first simply by being heavier at baseline, not because the treatment was effective.
-While the experimental design and protocol justify the assumption,
+While the experimental design and protocol justify the equal baseline assumption,
 its possible in practice inefficiencies in staging mice result in group level
@@
 result in the overall time to 1500mm^3^ to be different.
-
-Time to 1500mm^3^ is an artificial endpoint used for exploratory purposes in this analysis
+Time to 1500mm^3^ is also an artificial endpoint used for exploratory purposes in this analysis
 and future experiments could set more meaningful volume endpoints that
-may dramatically change these results.
+may change these results.
 
